My title

### 2024-12-03 paper/paper.qmd
finished variables section

The dataset used for this paper's analysis is titled "Canadian Grocery Price Data" and was obtained from Project Hammer [@projecthammer], which compiles grocery store price listings across the 8 major Canadian Grocers — Voila, T&T, Loblaws, No Frills, Metro, Galleria, Walmart, and Save-On-Foods. The dataset contains data starting from February 28, 2024 and is updated regularly with the addition of new entries. Only the entries up until November 25, 2024 are included within this paper's analysis.

@@

@@

The observations in the dataset are scraped from the grocer's official website's interface, which means it only contains the information that is publically listed and available. The recorded prices for each item are the listed price for the "in store pick up" option, with the target pickup neighbourhood being a neighbourhood in Toronto. Certain values and listings may be missed, since the internal APIs that power the grocer's websites are not accessed and specific extracts may error for certain vendors on particular days for unforeseeable reasons. Starting from July 11, the targetted variety of grocery items for data collection was greatly increased, meaning that there may be products which are missing pricing information before then.

## Data-Cleaning

After combining the two tables, `raw` and `product`, by matching each entry's product ID, the full dataset with $12,753,964$ entries was created. Then, entries with "egg" or "eggs", and "dozen" or "12" appearing in their product name or units were selected for. All February entries were omitted from analysis due to data collection starting on 28 February, 2024, which means there is a lack of sufficient data for that month to make any reliable conclusions from. The relevant variables of interest to us are: `nowtime`, `vendor`, `current_price`, `old_price`, and `product_name`, as explained below:

For ease of analysis, some additional variables were derived.

All entries with empty values in any of these key variables were dropped. It is important to note that this would filter out any unannounced price changes, as only price decreases for an explicitly advertised sale event would include both `current_price` and `old_price` variables. The first 5 rows from the cleaned dataset along with these derived variables is shown in @fig-datahead.

@@

## Outcome variables

Add graphs, tables and text. Use sub-sub-headings for each outcome variable or update the subheading to be singular.

```{r}
#| label: fig-planes
#| fig-cap: Relationship between wing length and width
#| echo: false
#| warning: false
#| message: false

#analysis_data |> 
#  ggplot(aes(x = width, y = length)) +
#  geom_point(alpha = 0.8) +
#  theme_minimal() +
#  labs(x = "Wing width (mm)",
#       y = "Wing length (mm)")
```

Talk way more about it.

## Predictor variables

Add graphs, tables and text.

Use sub-sub-headings for each outcome variable and feel free to combine a few into one if they go together naturally.

@@

# Additional data details {#sec-data-details}

!!! ADD FIRST 5 ROWS DATA HERE

@@

posterior_vs_prior(model1) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()
```

>>>>

The dataset used for this paper's analysis is titled "Canadian Grocery Price Data" and was obtained from Project Hammer [@projecthammer], which compiles grocery store price listings across the 8 major Canadian Grocers — Voila, T&T, Loblaws, No Frills, Metro, Galleria, Walmart, and Save-On-Foods. The dataset contains data starting from February 28, 2024 and is updated regularly with the addition of new entries. Only the entries up until November 25, 2024 are considered within this paper.

@@

The relevant variables of interest to us are: `nowtime`, `vendor`, `current_price`, `old_price`, and `product_name`, as explained below:

For ease of analysis, these additional variables were derived:

All entries with empty values in any of these key variables were dropped. Further details regarding the data cleaning process and a glimpse of the first 5 rows of the final cleaned dataset can be found in @sec-data-cleaning.

@@

The observations in the dataset are scraped from the grocer's official website's interface, which means it only contains the information that is publically listed and available. The recorded prices for each item are the listed price for the "in store pick up" option, with the target pickup neighbourhood being a neighbourhood in Toronto. Certain values and listings may be missed, since the internal APIs that power the grocer's websites are not accessed and specific extracts may error for certain vendors on particular days for unforeseeable reasons. Starting from July 11, the targeted variety of grocery items for data collection was greatly increased, meaning that there may be products which are missing pricing information before then.

## Variables of Interest

The outcome variable that we are interested in understanding is the `current_price`, representing the sale price of a dozen eggs. The predictor variables are `month`, `vendor`, `old_price`, and `prev_month_avg`.

@fig-vendorprices shows a boxplot of the sale pricing of a dozen eggs across the 8 vendors. From this graph, we can observe that T&T has the lowest prices overall and Save On Foods the highest. There are also differences in the range of prices offered, with T&T and Metro having much wider price ranges compared to Galleria and Save On Foods, which have highly consistent sale pricing.

```{r}
#| label: fig-vendorprices
#| fig-cap: Average Sale Price of a Dozen Eggs Across Vendors
#| echo: false
#| warning: false
#| message: false

analysis_data |> 
  ggplot(aes(x = vendor, y = current_price)) +
  geom_boxplot(outlier.colour="black", outlier.shape=16,
             outlier.size=2, notch=FALSE) +
  theme_classic() +
  labs(x = "Vendor",
       y = "Sale Price of a Dozen Eggs")
```

Next, @fig-monthprices shows the average sale price of a dozen eggs across different months. It can be observed that price dips slightly during the summer, reaching its lowest in July, before increasing again. Differences throughout months are typically gradual and do not fluctuate very much.

```{r}
#| label: fig-monthprices
#| fig-cap: Average Sale Price of a Dozen Eggs Across Months March to November
#| echo: false
#| warning: false
#| message: false

price_permonth <- analysis_data %>% 
  select(-prev_month_avg) %>%
  group_by(month) %>%
  summarize(avg_price = mean(current_price, na.rm = TRUE))

price_permonth <- price_permonth %>%
  mutate(month = recode(month,
                        Mar = 3,
                        Apr = 4,
                        May = 5,
                        Jun = 6,
                        Jul = 7,
                        Aug = 8, 
                        Sep = 9,
                        Oct = 10,
                        Nov = 11))

#locks in months in chronological order
price_permonth <- price_permonth[order(price_permonth$month, decreasing = FALSE),]

#converts back to month name while keeping order
price_permonth$month <- month.abb[price_permonth$month]
price_permonth$month <- factor(price_permonth$month, levels = price_permonth$month)

price_permonth |> 
  ggplot(mapping = aes(x = month, y = avg_price)) +
  geom_col() + 
  theme_classic() +
  labs(x = "Month",
       y = "Average Sale Price of a Dozen Eggs")
```

@fig-oldvsnew shows the positive and relatively linear relationship between the original presale price and discounted price of a dozen eggs across all 8 vendors. The dotted line has a slope of 1 and indicates how prices would look if discounted prices are equal to presale prices. All data points are on or below the dotted line, indicating that discounted prices are always equal to or less than the original presale price. The blue line, representing the line of best fit, has a positive but smaller slope than the dotted line, indicating that sale prices increase at a slower rate than presale prices. In other words, the greater an item's presale price, the greater the price difference would be. This confirms our intuitions about how sale events occur in the real world, as sales are typically calculated as a percentage of the original price.

```{r}
#| label: fig-oldvsnew
#| fig-cap: Presale and Sale Prices of a Dozen Eggs
#| echo: false
#| warning: false
#| message: false

analysis_data |> 
  ggplot(aes(x = old_price, y = current_price)) +
  geom_jitter() + 
  geom_smooth(method = lm, se = FALSE) +
  geom_abline(intercept = 0, slope = 1, show.legend = TRUE, linetype = "dashed") +
  theme_classic() +
  labs(x = "Presale Price",
       y = "Sale Price")

@@

@fig-prevmonth shows the relationship between the current sale price of a dozen eggs compared to the previous month's average price. Similar to @fig-oldvsnew, the dotted line has a slope of 1 and indicates variables growing at an equal rate whereas the blue line represents the line of best fit. Here, in contrast to @fig-oldvsnew, the relationship appears much weaker with a greater scatter and deviation of values. The line of best fit has a positive slope that is less than 1, indicating that current sale prices increase when previous month's prices do, but at a slower rate. That is, a higher price average in the previous month is correlated with higher prices in the current month, but not to a high degree.

```{r}
#| label: fig-prevmonth
#| fig-cap: Sale Price of a Dozen Eggs vs Previous Month's Average Price
#| echo: false
#| warning: false
#| message: false

analysis_data |> 
  ggplot(aes(x = prev_month_avg, y = current_price)) +
  geom_jitter(alpha = 0.5) +
  geom_smooth(method = lm, se = FALSE) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
  theme_classic() +
  labs(x = "Previous Month's Average Price of a Dozen Eggs",
       y = "Sale Price")

## Other Datasets

There are other food pricing related datasets available for use, such as Statistics Canada's "Average Retail Food Prices" [@otherdataset]. Though it does also contain average price information for a dozen eggs, this dataset was not utilized within this paper's analysis in the end because it only contains monthly averages that are also not divided by vendor. As well, the averages in this dataset are generalized across the entire provinces, in contrast to Project Hammer's dataset which is centred on Toronto prices [@projecthammer]. This could introduce inaccuracies into our model, as the wider geographical range may be failing to capture Toronto's comparatively higher cost of living, compared to other Ontario regions.

@@

# Additional data details

## Data-Cleaning {#sec-data-cleaning}

After combining the two tables, `raw` and `product`, by matching each entry's product ID, the full dataset with $12,753,964$ entries was created. Then, entries with "egg" or "eggs", and "dozen" or "12" appearing in their product name or units were selected for. All February entries were omitted from analysis due to data collection starting on 28 February, 2024, which means there is a lack of sufficient data for that month to make any reliable conclusions from.

All unannounced price changes, including increases or decreases, have only a `current_price` value and empty `old_price` values, so all such entries were filtered out to narrow down analysis onto explicitly advertised sale events only. Then, the `current_price` and `old_price` columns were filtered to ensure all values were non-negative, as these entries are evidently the result of measurement error during data scraping.

After cleaning the dataset and adding in the derived variables, the final dataset of $1,278$ entries is what analysis will be conducted on. The first 5 rows from this analysis data is shown in @fig-datahead.

@@

posterior_vs_prior(model1) +
  theme_classic() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()
```
